---
title: 'GO of the gene clusters in the physical space'
author: "Daniel Paysan"
date: "06/07/2022"
output: html_document
---

This notebook conducts a GO analysis to characterize the gene clusters identified in the physical space.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/home/daniel/PycharmProjects/image2reg/")
```

# 0. Environmental setup

First, we load the required packages.

```{r environmental setup}
require("topGO")
require("KEGGREST")
require("org.Hs.eg.db")
require("GeneAnswers")
require("EnrichmentBrowser")
```
Next, we load a function that enables a simple GO analysis using the topGO API.

```{r}
# taken from https://github.com/karthikshekhar/CellTypeMIMB/blob/master/utilities.R
topGOterms = function(fg.genes = NULL,
                      bg.genes = NULL,
                      organism = "Mouse",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 100) {
  if (is.null(fg.genes) | is.null(bg.genes)) {
    stop("Error : Both gene lists are empty")
  }
  
  require(topGO)
  if (organism == "Mouse") {
    mapping.use = "org.Mm.eg.db"
    require(org.Mm.eg.db)
  } else if (organism == "Human") {
    mapping.use = "org.Hs.eg.db"
    require(org.Hs.eg.db)
  } else {
    stop("Error : Organisms other than mouse not supported currently")
  }
  
  n = length(bg.genes)
  geneList = integer(n)
  names(geneList) = bg.genes
  geneList[intersect(names(geneList), fg.genes)] = 1
  print(paste0(
    "Total ",
    length(geneList),
    " genes. ",
    sum(geneList),
    " genes in the foreground"
  ))
  geneList = factor(geneList)
  
  if (ontology.use %in% c("BP", "CC", "MF")) {
    print(paste0("Using Ontology : ", ontology.use))
  } else {
    stop("Error: Ontology not available. Should be one of BP, CC or MF")
  }
  # Make GO object
  GOdata <- new(
    "topGOdata",
    description = "GOanalysis",
    ontology = ontology.use,
    allGenes = geneList,
    annot = annFUN.org,
    mapping = mapping.use,
    ID = "SYMBOL",
    nodeSize = 10
  )
  print(paste0(
    "Using the ",
    stats.use,
    " statistic with the ",
    algorithm.use,
    " algorithm"
  ))
  res.result <-
    runTest(GOdata, statistic = stats.use, algorithm = algorithm.use)
  to.return = list()
  to.return$GOdata = GOdata
  
  allGO = usedGO(object = GOdata) 
  
  to.return$res.table <-
    GenTable(GOdata,
             pval = res.result,
             topNodes = length(allGO),
             numChar = num.char)
  return(to.return)
}
```


# 1. Gene Ontology analyses

We now conduct the gene ontology analysis for the individual clusters and correct the computed p-values for multiple testing using the Bonferroni-Holm procedure.
To characterize each cluster, we look at the top-10 GO terms.


The background is given by all identified physical regulator genes.

```{r}
all_genes <- c('CEBPA', 'CXXC4', 'E2F1', 'ERG', 'JUN', 'FGFR3', 'MAP3K5', 'MAPK7', 'MOS', 'PIK3R2', 'PRKCE', 'RELB', 'SRC', 'TRAF2', 'WWTR1', 'ATF4', 'BAX', 'CREB1', 'ELK1', 'IRAK4', 'SMAD4', 'TSC2', 'AKT1S1', 'BRAF', 'CDC42', 'DIABLO', 'FOXO1', 'GLI1', 'MAP2K3', 'MAP3K2', 'MAP3K9', 'MYD88', 'PRKACA', 'RAF1', 'RHOA', 'SMO', 'SREBF1', 'CDKN1A', 'BCL2L11', 'CASP8', 'HRAS')
```


```{r}
selected_genes <- c('SREBF1', 'WWTR1', 'FOXO1', 'DIABLO', 'MYD88', 'RHOA', 'SMO')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```

```{r}
selected_genes <- c('PIK3R2', 'TRAF2', 'FGFR3', 'AKT1S1', 'PRKCE', 'MOS', 'SRC')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```

```{r}
selected_genes <- c('BRAF', 'RAF1', 'CDC42', 'MAP3K2', 'MAP3K9')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```

```{r}
selected_genes <- c('BCL2L11', 'CASP8')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```

```{r}
selected_genes <- c('RELB', 'MAP3K5', 'MAP2K3', 'MAPK7')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```

```{r}
selected_genes <- c('IRAK4', 'BAX', 'SMAD4', 'TSC2', 'CREB1', 'ATF4')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```

```{r}
selected_genes <- c('CEBPA', 'JUN', 'ERG')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```

```{r}
selected_genes <- c('ELK1', 'E2F1')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```

```{r}
selected_genes <- c('PRKACA', 'CXXC4', 'GLI1')
steiner_go_results <-  topGOterms(fg.genes = selected_genes,
                      bg.genes = all_genes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
steiner_go_results$res.table[,"FDR"] <- p.adjust(steiner_go_results$res.table$pval, method="fdr")
head(steiner_go_results$res.table, 10)
```