---
title: 'Enrichment analysis of the gene set of in the inferred gene-gene interactome'
author: "Daniel Paysan"
date: "06/07/2022"
output: html_document
---

This notebook conducts a GO analysis to characterize the gene set identified as part of the gene-gene interactome output by the prize-collecting Steiner network analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/media/paysan_d/T7/image2reg")
```

# 0. Environmental setup

First, we load the required packages.

```{r environmental setup}
require("topGO")
require("KEGGREST")
require("org.Hs.eg.db")
require("GeneAnswers")
require("EnrichmentBrowser")
require("ggplot2")
require("stringr")
```
Next, we load a function that enables a simple GO analysis using the topGO API.

```{r}
# taken from https://github.com/karthikshekhar/CellTypeMIMB/blob/master/utilities.R
topGOterms = function(fg.genes = NULL,
                      bg.genes = NULL,
                      organism = "Mouse",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 100) {
  if (is.null(fg.genes) | is.null(bg.genes)) {
    stop("Error : Both gene lists are empty")
  }
  
  require(topGO)
  if (organism == "Mouse") {
    mapping.use = "org.Mm.eg.db"
    require(org.Mm.eg.db)
  } else if (organism == "Human") {
    mapping.use = "org.Hs.eg.db"
    require(org.Hs.eg.db)
  } else {
    stop("Error : Organisms other than mouse not supported currently")
  }
  
  n = length(bg.genes)
  geneList = integer(n)
  names(geneList) = bg.genes
  geneList[intersect(names(geneList), fg.genes)] = 1
  print(paste0(
    "Total ",
    length(geneList),
    " genes. ",
    sum(geneList),
    " genes in the foreground"
  ))
  geneList = factor(geneList)
  
  if (ontology.use %in% c("BP", "CC", "MF")) {
    print(paste0("Using Ontology : ", ontology.use))
  } else {
    stop("Error: Ontology not available. Should be one of BP, CC or MF")
  }
  # Make GO object
  GOdata <- new(
    "topGOdata",
    description = "GOanalysis",
    ontology = ontology.use,
    allGenes = geneList,
    annot = annFUN.org,
    mapping = mapping.use,
    ID = "SYMBOL",
    nodeSize = 10
  )
  print(paste0(
    "Using the ",
    stats.use,
    " statistic with the ",
    algorithm.use,
    " algorithm"
  ))
  res.result <-
    runTest(GOdata, statistic = stats.use, algorithm = algorithm.use)
  to.return = list()
  to.return$GOdata = GOdata
  
  allGO = usedGO(object = GOdata) 
  
  to.return$res.table <-
    GenTable(GOdata,
             pval = res.result,
             topNodes = length(allGO),
             numChar = num.char)
  return(to.return)
}
```


```{r}
barplot_topgo <- function(data){
  p <- ggplot(data, aes(x=Term, y=-log10(qval))) + 
    geom_bar(stat="identity", fill = "#FF6666") + 
    scale_x_discrete(labels = function(x) str_wrap(x, width = 30))+
    geom_hline(yintercept = c(-log10(0.05)), linetype = c("dashed"), colour = c("black"), size = c(0.5))+
  theme_bw(base_size = 10) +
  theme(
    legend.position = 'right',
    legend.background = element_rect(),
    plot.title = element_text(angle = 0, size = 16, face = 'bold', vjust = 1),
    plot.subtitle = element_text(angle = 0, size = 10, face = 'bold', vjust = 1),
    plot.caption = element_text(angle = 0, size = 10, face = 'bold', vjust = 1),

    axis.text.x = element_text(angle = 90, size = 10, face = 'bold', hjust = 0.5, vjust=0.5),
    axis.text.y = element_text(angle = 0, size = 10, face = 'bold', vjust = 0.5),
    axis.title = element_text(size = 10, face = 'bold'),
    axis.title.x = element_text(size = 10, face = 'bold'),
    axis.title.y = element_text(size = 10, face = 'bold'),
    axis.line = element_line(colour = 'black'),

    #Legend
    legend.key = element_blank(), # removes the border
    legend.key.size = unit(1, "cm"), # Sets overall area/size of the legend
    legend.text = element_text(size = 10, face = "bold"), # Text size
    title = element_text(size = 14, face = "bold")) 
  #+coord_flip()
  return(p)
}
```


# 1. Gene Ontology analysis

To characterize the importance of the selected subset of genes in the inferred GGI we conduct a gene ontology analysis testing for the enrichment of terms associated with a biological process using Fisher's exact test and correct for multiple testing using the Bonferroni-Holm procedure. Therefore, we first load the information into memory which nodes were selected by the Prize-collecting Steiner tree analysis.

```{r}
node_selection <- read.csv("data/experiments/rohban/interactome/inference_results/node_selection.csv", row.names=1)

ppi_nodes <- rownames(node_selection)
ggi_nodes <- rownames(node_selection)[node_selection$selected_in_ggi==1]
```




```{r}
ggi_go_results <-  topGOterms(fg.genes = ggi_nodes,
                      bg.genes = ppi_nodes,
                      organism = "Human",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 70)
ggi_go_results$res.table[,"qval"] <- p.adjust(ggi_go_results$res.table$pval, method="BH")
head(ggi_go_results$res.table, 20)
```


```{r, fig.height=6, fig.width=14}
sig_ggi_go_terms <- ggi_go_results$res.table[1:30,]
sig_ggi_go_terms$pval <- as.numeric(sig_ggi_go_terms$pval)
p <- barplot_topgo(sig_ggi_go_terms)
p + ggtitle("Top-30 GO terms for the GGI", subtitle = "249 selected genes") + 
  labs(x="GO term", y="-log10(adjusted p value)") + scale_y_continuous(breaks=c(0:10))
```
```



# 2. Discussion

The previous analysis shows that 7 of the 9 non-singleton clusters are significantly enriched for at least one GO term associated to a biological processes when considering the raw p-values obtained from the topGO analysis which are assumed to be not affected by multiple testing.