---
title: "GO analysis of the gene embedding clusters"
output: html_notebook
---

This notebook assess the enrichment of the individual clusters of the inferred gene embeddings for specific pathways.

# 0. Environmental setup

Please update the file path for root.dir argument to the location of the image2reg repository.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/home/paysan_d/PycharmProjects/image2reg/")
```

Please skip the next chunk if the packages are already installed.
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager", version = "3.11")

BiocManager::install("clusterProfiler")
BiocManager::install("org.Hs.eg.db")
install.packages("tidyverse")


```



```{r}
require("clusterProfiler")
require("org.Hs.eg.db")
require("ggplot2")
require("stringr")
```

Please verify that you have R version 4.0.3 installed and clusterProfiler version 3.18.1. and GO.db version 3.12.1.
This is important to reproduce the results of the paper as the definition of gene sets uses in the GO analyses is subject to constant change.
To ensure the reproducibility of our results make sure that the package version numbers of in particular clusterProfiler match these printed as the output of the cell below.
The versions are also listed in the "gene_embedding_cluster_analyses.nb.html" file located in the same directory as this script.

Finally, we also provide the R environment with the results obtained for the specified configuration in form of the "other/gene_embedding_cluster_data.Rdata" file, which can be loaded by removing the preceding "#" from the code chunk on line 221.

```{r}
sessionInfo()
```


```{r}
# taken from https://github.com/karthikshekhar/CellTypeMIMB/blob/master/utilities.R
topGOterms = function(fg.genes = NULL,
                      bg.genes = NULL,
                      organism = "Mouse",
                      ontology.use = "BP",
                      stats.use = "fisher",
                      algorithm.use = "weight01",
                      num.char = 100) {
  if (is.null(fg.genes) | is.null(bg.genes)) {
    stop("Error : Both gene lists are empty")
  }
  
  require(topGO)
  if (organism == "Mouse") {
    mapping.use = "org.Mm.eg.db"
    require(org.Mm.eg.db)
  } else if (organism == "Human") {
    mapping.use = "org.Hs.eg.db"
    require(org.Hs.eg.db)
  } else {
    stop("Error : Organisms other than mouse not supported currently")
  }
  
  n = length(bg.genes)
  geneList = integer(n)
  names(geneList) = bg.genes
  geneList[intersect(names(geneList), fg.genes)] = 1
  print(paste0(
    "Total ",
    length(geneList),
    " genes. ",
    sum(geneList),
    " genes in the foreground"
  ))
  geneList = factor(geneList)
  
  if (ontology.use %in% c("BP", "CC", "MF")) {
    print(paste0("Using Ontology : ", ontology.use))
  } else {
    stop("Error: Ontology not available. Should be one of BP, CC or MF")
  }
  # Make GO object
  GOdata <- new(
    "topGOdata",
    description = "GOanalysis",
    ontology = ontology.use,
    allGenes = geneList,
    annot = annFUN.org,
    mapping = mapping.use,
    ID = "SYMBOL",
    nodeSize = 10
  )
  print(paste0(
    "Using the ",
    stats.use,
    " statistic with the ",
    algorithm.use,
    " algorithm"
  ))
  res.result <-
    runTest(GOdata, statistic = stats.use, algorithm = algorithm.use)
  to.return = list()
  to.return$GOdata = GOdata
  
  allGO = usedGO(object = GOdata) 
  
  to.return$res.table <-
    GenTable(GOdata,
             pval = res.result,
             topNodes = length(allGO),
             numChar = num.char)
  return(to.return)
}
```


```{r}
barplot_topgo <- function(data){
  p <- ggplot(data[1:5,], aes(x=Description, y=-log10(p.adjust))) + 
    geom_bar(stat="identity", fill = "#FF6666") + 
    scale_x_discrete(labels = function(x) str_wrap(x, width = 25))+
    geom_hline(yintercept = c(-log10(0.05)),
    linetype = c("dashed"),
    colour = c("black"),
    size = c(0.5))+
  theme_bw(base_size = 10) +
  theme(
    legend.position = 'right',
    legend.background = element_rect(),
    plot.title = element_text(angle = 0, size = 16, face = 'bold', vjust = 1),
    plot.subtitle = element_text(angle = 0, size = 10, face = 'bold', vjust = 1),
    plot.caption = element_text(angle = 0, size = 10, face = 'bold', vjust = 1),

    axis.text.x = element_text(angle = 0, size = 10, face = 'bold', hjust = 1.10),
    axis.text.y = element_text(angle = 0, size = 10, face = 'bold', vjust = 0.5),
    axis.title = element_text(size = 10, face = 'bold'),
    axis.title.x = element_text(size = 10, face = 'bold'),
    axis.title.y = element_text(size = 10, face = 'bold'),
    axis.line = element_line(colour = 'black'),

    #Legend
    legend.key = element_blank(), # removes the border
    legend.key.size = unit(1, "cm"), # Sets overall area/size of the legend
    legend.text = element_text(size = 10, face = "bold"), # Text size
    title = element_text(size = 14, face = "bold")) +
  coord_flip()
  return(p)
}
```

```{r}
barplot_topgo_wo_dashed <- function(data){
  p <- ggplot(data[1:5,], aes(x=Description, y=-log10(p.adjust))) + 
    geom_bar(stat="identity", fill = "#FF6666") + 
    scale_x_discrete(labels = function(x) str_wrap(x, width = 25))+
  theme_bw(base_size = 10) +
  theme(
    legend.position = 'right',
    legend.background = element_rect(),
    plot.title = element_text(angle = 0, size = 16, face = 'bold', vjust = 1),
    plot.subtitle = element_text(angle = 0, size = 10, face = 'bold', vjust = 1),
    plot.caption = element_text(angle = 0, size = 10, face = 'bold', vjust = 1),

    axis.text.x = element_text(angle = 0, size = 10, face = 'bold', hjust = 1.10),
    axis.text.y = element_text(angle = 0, size = 10, face = 'bold', vjust = 0.5),
    axis.title = element_text(size = 10, face = 'bold'),
    axis.title.x = element_text(size = 10, face = 'bold'),
    axis.title.y = element_text(size = 10, face = 'bold'),
    axis.line = element_line(colour = 'black'),

    #Legend
    legend.key = element_blank(), # removes the border
    legend.key.size = unit(1, "cm"), # Sets overall area/size of the legend
    legend.text = element_text(size = 10, face = "bold"), # Text size
    title = element_text(size = 14, face = "bold")) +
  coord_flip()
  return(p)
}
```


# 1. Read in data

```{r}
cluster_data <- read.csv("data/experiments/rohban/interactome/cluster_infos/all_gene_embeddings_clusters.csv", row.names = 1)
```

# 2. Enrichment analysis

```{r}
universe <- cluster_data$gene
universe<-bitr(universe, fromType = "SYMBOL", toType = "ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
universe <- sort(universe, decreasing=T)
```
```{r}
gene_clusters <- list()
for (i in sort(unique(cluster_data$cluster))){
  gene_clusters[[as.character(i)]] <- cluster_data[cluster_data$cluster == i, "gene"]
}
```

```{r}
for(k in names(gene_clusters)){
  gene_clusters[[k]] <- bitr(gene_clusters[[k]], fromType = "SYMBOL", toType = "ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
}
```

```{r, fig.width=10, fig.heigh=20}
compareGO <- compareCluster(geneClusters=gene_clusters, fun="enrichGO", data="", OrgDb="org.Hs.eg.db", pAdjustMethod="fdr", ont="BP", universe=universe, pvalueCutoff=0.05, qvalueCutoff=10,  maxGSSize = 500, minGSSize = 10)
cluster_results = compareGO@compareClusterResult
```

To speed up the computations, one can also directly load the results of the GO analysis.

```{r}
#load("other/gene_embedding_cluster_data.Rdata")
```


```{r, fig.width=9, fig.height=8}
dotplot(compareGO, font.size=15, showCategory=3,) + theme(
  axis.text.y = element_text(face="bold", 
                             size=15))
```


## Enrichment cluster 1

```{r}
compareGO_noco <- compareCluster(geneClusters=gene_clusters, fun="enrichGO", data="", OrgDb="org.Hs.eg.db", pAdjustMethod="fdr", ont="BP", universe=universe, pvalueCutoff=1, qvalueCutoff=10,  maxGSSize = 500, minGSSize = 10)
cluster_results_noco = compareGO_noco@compareClusterResult
```


```{r, fig.height=4, fig.width=6}
sig_cluster0_go_terms <- cluster_results_noco[cluster_results_noco$Cluster ==0,]
sig_cluster0_go_terms$pval <- as.numeric(sig_cluster0_go_terms$pval)
p <- barplot_topgo(sig_cluster0_go_terms)
p + ggtitle("GO Terms for Cluster 1", subtitle = "") + 
  labs(x="GO term", y="-log10(FDR)")
```

## Enrichment Cluster 2

```{r, fig.height=4, fig.width=6}
sig_cluster0_go_terms <- cluster_results_noco[cluster_results_noco$Cluster ==1,]
sig_cluster0_go_terms$pval <- as.numeric(sig_cluster0_go_terms$pval)
p <- barplot_topgo(sig_cluster0_go_terms)
p + ggtitle("GO Terms for Cluster 2", subtitle = "") + 
  labs(x="GO term", y="-log10(FDR)")
```
## Enrichment Cluster 3

```{r, fig.height=4, fig.width=6}
sig_cluster0_go_terms <- cluster_results_noco[cluster_results_noco$Cluster ==2,]
sig_cluster0_go_terms$pval <- as.numeric(sig_cluster0_go_terms$pval)
p <- barplot_topgo(sig_cluster0_go_terms)
p + ggtitle("GO Terms for Cluster 3", subtitle = "") + 
  labs(x="GO term", y="-log10(FDR)")
```
# Enrichment of Cluster 4

```{r, fig.height=4, fig.width=6}
sig_cluster0_go_terms <- cluster_results_noco[cluster_results_noco$Cluster ==3,]
sig_cluster0_go_terms$pval <- as.numeric(sig_cluster0_go_terms$pval)
p <- barplot_topgo(sig_cluster0_go_terms)
p + ggtitle("GO Terms for Cluster 4", subtitle = "") + 
  labs(x="GO term", y="-log10(FDR)")
```

## Enirchment of Cluster 5

```{r, fig.height=4, fig.width=6}
sig_cluster0_go_terms <- cluster_results_noco[cluster_results_noco$Cluster ==4,]
sig_cluster0_go_terms$pval <- as.numeric(sig_cluster0_go_terms$pval)
p <- barplot_topgo_wo_dashed(sig_cluster0_go_terms)
p + ggtitle("GO Terms for Cluster 5", subtitle = "") + 
  labs(x="GO term", y="-log10(FDR)")
```

## Enrichment Cluster 6

```{r, fig.height=4, fig.width=6}
sig_cluster0_go_terms <- cluster_results_noco[cluster_results_noco$Cluster ==5,]
sig_cluster0_go_terms$pval <- as.numeric(sig_cluster0_go_terms$pval)
p <- barplot_topgo(sig_cluster0_go_terms)
p + ggtitle("GO Terms for Cluster 6", subtitle = "") + 
  labs(x="GO term", y="-log10(FDR)")
```
# Enrichment Cluster 7

```{r, fig.height=4, fig.width=6}
sig_cluster0_go_terms <- cluster_results_noco[cluster_results_noco$Cluster ==6,]
sig_cluster0_go_terms$pval <- as.numeric(sig_cluster0_go_terms$pval)
p <- barplot_topgo(sig_cluster0_go_terms)
p + ggtitle("GO Terms for Cluster 7", subtitle = "") + 
  labs(x="GO term", y="-log10(FDR)")
```

## Enrichment for Cluster 8

```{r, fig.height=4, fig.width=6}
sig_cluster0_go_terms <- cluster_results_noco[cluster_results_noco$Cluster ==7,]
sig_cluster0_go_terms$pval <- as.numeric(sig_cluster0_go_terms$pval)
p <- barplot_topgo(sig_cluster0_go_terms)
p + ggtitle("GO Terms for Cluster 8", subtitle = "") + 
  labs(x="GO term", y="-log10(FDR)")
```

```{r}
#save.image(file = "other/gene_embedding_cluster_data.Rdata")
sessionInfo()
```

